\section{Coordination Capacity}

This section formalizes the notion of \emph{coordination capacity} for open,
adversarial monetary systems, isolates irreversibility as the primitive
resource, and states the U₁ Coordination Capacity Conjecture.


\subsection{Irreversibility as the Primitive}

\begin{definition}[Irreversible Work Step]
A work step is \emph{irreversible} if its physical realization increases
entropy in the environment such that no observer with bounded resources
can algorithmically reconstruct a pre-image of the microscopic state
prior to the step.

We write $E_{\irr} > 0$ for the minimal energy cost of such a step at
a given ambient temperature.
\end{definition}

\begin{definition}[Physics-Anchored Protocol]
A protocol $\Pi$ is \emph{physics-anchored} if every admissible block
production event must include at least one irreversible work step
whose successful completion is:
\begin{enumerate}
  \item publicly and efficiently verifiable, and
  \item infeasible to simulate without incurring comparable
        physical energy cost.
\end{enumerate}
\end{definition}

In other words, physics-anchored protocols encode consensus progress
into the irreversibility of the underlying physical process (e.g. PoW),
rather than into institutional decree or access-controlled identities.


\subsection{Atomic Unit of Account}

\begin{definition}[Atomic Unit of Account]
Let $\Pi$ be a physics-anchored monetary protocol with total supply
process $(U_t)_{t \ge 0}$.

An \emph{atomic unit of account} $U$ is the smallest denomination such that:
\begin{enumerate}
  \item (Indivisibility) No valid ledger state can represent a claim
        smaller than $U$.
  \item (Fungibility) Any two claims of size $U$ are mutually
        substitutable under the protocol's transfer rules.
  \item (Monetary Closure) The aggregate supply satisfies
        $\sum U_t \le S_{\max}$ across all valid histories.
  \item (Settlement Granularity) Every irreversible settlement
        event can be decomposed into a finite change in the allocation
        of atomic units $U$.
\end{enumerate}
\end{definition}

\begin{remark}
In Bitcoin, the atomic unit is $U = 1$ satoshi and $S_{\max} = 21 \times
10^6 \text{ BTC} = 2.1 \times 10^{15}$ satoshis. For any protocol
satisfying A₄ (Monetary Closure), an analogous $U$ exists.
\end{remark}


\subsection{Coordination Flow and Capacity}

We now formalize coordination as an information flow from latent
economic reality into a publicly settled ledger.

\begin{definition}[Economic State and Ledger State]
Let $(\Theta_t)_{t \ge 0}$ denote a (possibly high-dimensional)
stochastic process capturing the true economic preferences and claims
of agents at time $t$, and let $(A_t)_{t \ge 0}$ denote the publicly
settled ledger state produced by protocol $\Pi$ at time $t$.

Both processes are defined on a common probability space, and
$(A_t)$ is adapted to the filtration generated by protocol-visible
messages and blocks.
\end{definition}

\begin{definition}[Coordination Flow]
For a fixed protocol $\Pi$, the \emph{coordination flow rate} is
\[
\dot V_{\coord}(\Pi)
=
\lim_{T \to \infty}
\frac{1}{T}\, I(\Theta_{0:T}; A_{0:T}),
\]
whenever the limit exists, where $I(\cdot;\cdot)$ denotes mutual
information.
\end{definition}

Intuitively, $\dot V_{\coord}(\Pi)$ measures how many bits of true
economic intent are irreversibly compressed into the public ledger per
unit time.

\begin{definition}[Coordination Capacity]
Let $\mathcal{M}_{\phys}$ denote the class of physics-anchored protocols
satisfying:
\begin{enumerate}
  \item public verifiability (A₂),
  \item open participation (permissionless entry and exit),
  \item bounded adversary resources (e.g.\ an honest-majority work budget),
  \item physically irreversible settlement cost (A₁).
\end{enumerate}

The \emph{coordination capacity} of the environment is
\[
C_{\coord}
=
\sup_{\Pi \in \mathcal{M}_{\phys}}
\dot V_{\coord}(\Pi)
=
\sup_{\Pi \in \mathcal{M}_{\phys}}
\lim_{T \to \infty}
\frac{1}{T}\, I(\Theta_{0:T}; A_{0:T}).
\]
\end{definition}

Thus $C_{\coord}$ is to adversarial economic settlement what Shannon
capacity is to communication over noisy channels: the maximal
information rate at which true economic state can be mapped into a
canonical public ledger with arbitrarily small corruption probability,
under fixed physical and adversarial constraints.


\subsection{Security Exponent and Irreversibility Density}

We now connect settlement reliability to the rate of irreversible work
performed by honest participants.

\begin{definition}[Security Exponent]
Consider a physics-anchored protocol $\Pi$ exhibiting probabilistic
finality (A₃). For a reorganization depth parameter $k \in \mathbb{N}$,
let
\[
P_{\reorg}(k) = \Pr(\text{ledger reorganization of depth} \ge k).
\]

If there exists $\beta > 0$ and $k_0$ such that
\[
P_{\reorg}(k) \le \exp(-\beta k)
\quad
\text{for all } k \ge k_0,
\]
we say that $\Pi$ has \emph{security exponent} $\beta$.
\end{definition}

\begin{definition}[Irreversibility Density]
Let $\dot E_{\honest}$ denote the long-run honest irreversible work
rate (in joules per second) invested in maintaining consensus for
$\Pi$, and let $\dot U_{\secured}$ denote the long-run rate (in units
per second) at which atomic units $U$ are involved in economically
meaningful settled transfers.

The \emph{irreversibility density} of $\Pi$ is
\[
\rho_{\irr}(\Pi)
=
\frac{\dot E_{\honest}}{\dot U_{\secured}}
\quad
\text{with units}
\quad
\frac{\mathrm{J}}{\mathrm{U}}.
\]
\end{definition}

\begin{remark}
Heuristically, $\rho_{\irr}$ can be interpreted as the average amount
of physically irreversible work crystallized into the ledger per unit
of account. Systems with $\rho_{\irr} \to 0$ cannot maintain a
meaningful barrier against adversarial rewriting of history in an
open, permissionless setting.
\end{remark}

In PoW-style U₁ systems, empirical and theoretical arguments suggest
that, under competitive mining and difficulty adjustment (T₁, T₂),
there exists a protocol-dependent constant $c > 0$ such that
\[
\beta(\Pi) \approx c \cdot
\frac{\dot E_{\honest}}{V_{\secured}},
\]
where $V_{\secured}$ is the total economic value (e.g.\ in units $U$
times purchasing power) that depends on $\Pi$ for settlement.\footnote{
The precise choice of $V_{\secured}$ (e.g.\ fee-paying volume, non-churn
flow, or broader balance at risk) is an empirical modeling decision.
For the purposes of the conjecture, it suffices that $V_{\secured}$
captures the value whose coordination depends critically on $\Pi$.}


\subsection{Model Boundary: Physics vs.\ Coercion}

The above definitions explicitly restrict attention to
\emph{physics-anchored} protocols in $\mathcal{M}_{\phys}$.

\begin{definition}[Institution-Anchored Ledger]
A ledger $\Pi_{\inst}$ is \emph{institution-anchored} if irreversibility
of entries is enforced primarily by:
\begin{enumerate}
  \item legal decree or administrative authority,
  \item access-controlled identities or whitelists,
  \item coercive enforcement (e.g.\ courts, police, capital controls),
\end{enumerate}
rather than by publicly verifiable irreversible work steps.
\end{definition}

Fiat currencies, bank databases, and permissioned ledgers typically
fall into this class: they may exhibit practical irreversibility within
a jurisdiction, but it is not rooted in physics and is not
permissionless or globally auditable in the sense of A₁–A₄.

\begin{remark}
Coordination capacity $C_{\coord}$ is defined only over
$\mathcal{M}_{\phys}$. Institution-anchored systems may coordinate
agents via coercive trust hierarchies, but they do not approach a
physics-limited capacity in the Shannon sense, and they do not provide
irreversibility that is independent of specific institutional
arrangements.
\end{remark}


\subsection{U₁ Coordination Capacity Conjecture}

We can now restate the core conjecture in this refined language.

\begin{conjecture}[U₁ Coordination Capacity Conjecture]
Consider the class $\mathcal{M}_{\phys}$ of physics-anchored protocols
operating in an open, permissionless, adversarial environment with a
bounded-work attacker (honest-majority assumption).

Then:
\begin{enumerate}
  \item A protocol $\Pi \in \mathcal{M}_{\phys}$ achieves strictly
        positive coordination capacity ($\dot V_{\coord}(\Pi) > 0$)
        if and only if it satisfies the U₁ axioms:
        A₁ (energy-bounded issuance),
        A₂ (public verifiability),
        A₃ (probabilistic finality),
        A₄ (monetary closure).

  \item For any such $\Pi$ with adversary work fraction $q < p$
        (honest fraction $p$), there exists $\beta(\Pi) > 0$ such that
        the reorganization probability obeys an exponential bound
        \[
        \Pr(\mathrm{reorg\ depth} \ge k)
        \;\le\;
        \left(\frac{q}{p}\right)^k
        =
        \exp(-\beta(\Pi)\, k)
        \quad \text{for all sufficiently large } k.
        \]

  \item Among protocols satisfying A₁–A₄, U₁-style PoW monetary systems
        can achieve security exponents $\beta(\Pi)$ arbitrarily close
        to the physical limit imposed by their irreversibility density:
        \[
        \beta(\Pi)
        \lesssim
        c \cdot
        \frac{\dot E_{\honest}}{V_{\secured}},
        \]
        for some protocol- and environment-dependent constant
        $c > 0$, with equality approached in the limit of optimal
        encoding of economic state into ledger state.
\end{enumerate}
\end{conjecture}

\begin{corollary}[Vanishing Irreversibility Density]
Within $\mathcal{M}_{\phys}$, any family of protocols
$(\Pi_n)_{n \in \mathbb{N}}$ with irreversibility density
$\rho_{\irr}(\Pi_n) \to 0$ as $n \to \infty$ satisfies
\[
\lim_{n \to \infty} \dot V_{\coord}(\Pi_n) = 0.
\]
In particular, physics-anchored systems whose marginal unit of account
is not backed by a nontrivial amount of irreversible work cannot
achieve nonzero coordination capacity in the limit.
\end{corollary}

The conjecture thus states that U₁ systems are to economic coordination
what Shannon-capacity-achieving codes are to communication: they
saturate a physically imposed limit on reliable performance in a noisy,
adversarial universe.